Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers Across Aurignacian Time Periods

### 2025-03-04 endofquarterpaper.qmd
edits

Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the sub-periods of the Aurignacian period. The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans. Chronology of the Aurignacian is primarly determined by correspondance with those material culture markers, which include carinated scrapers, small-flake tool production, and split-based bone and ivory tools [@Chu_Richter_2020]. The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015]. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k cal BP, and the early Aurignacian at 39.8 - 37.8 k cal BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ended with Greenland Interstadial 8.

<!-- add specific details about cultural markers, both for aurignacian and for each time period (create table in excel with name, date range, artifact indicators, environmental context, sources in the caption)-->

```{r}
library(knitr)

[@Jöris_Street_2008, @Banks_dErrico_Zilhao_2013, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021, @Sanz-Royo_Terlato_Marín-Arroyo_2024, @Chu_Richter_2020], 
```

In this study, we aim to address the following questions: How does sign distribution change across phases of the Aurignacian? How do the groupings in sign distribution vary in diversity, size, connectivity, and boundaries? What do the patterns tell us about dynamics of social organization throughout the Aurignacian? 
Firstly, we identify clusters within each time period using seriation and network analysis. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.

@@

```{r}
## clean the data
library(tidyverse)
library(magrittr)
library(rcarbon)
library(readxl)
signbase_full <- read_csv("data/signBase_Version1.0.csv")
nsites <- signbase_full %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts <- nrow(signbase_full)
ncountries <- signbase_full %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes <- ncol(signbase_full)

@@

signbase_full_clean <- signbase_full_clean %>%
  mutate(time_period = case_when((MedianBP > 42999) ~ "transitional",
                                 (MedianBP > 39899) ~ "proto_aurignacian",
                                 (MedianBP > 36999) ~ "early_aurignacian",
                                 (MedianBP > 31999) ~ "evolved_aurignacian"))

@@

trans_unique_data <- trans_unique_data %>% 
  mutate(group = case_when((site_name == "Pod Hradem") ~ "1",
                           (site_name == "El Castillo" | site_name == "Hohle Fels") ~ "2"))

proto_unique_data <- proto_unique_data %>% 
  mutate(group = case_when((site_name == "Labeko Koba" | 
                              site_name == "Gatzarria" | 
                              site_name == "Hohlenstein-Stadel" | 
                              site_name == "Mladeč" | 
                              site_name == "Abri Pataud" | 
                              site_name == "Fumane") ~ "1",
                           (site_name == "Geissenklösterle" |
                              site_name == "Grotte du Renne" |
                              site_name == "Vogelherd" |
                              site_name == "Spy" |
                              site_name == "La Ferrassie" |
                              site_name == "Les Cottés") ~ "2"))

@@

evolved_unique_data <- evolved_unique_data %>% 
  mutate(group = case_when((site_name == "Les Rois" |
                              site_name == "Gargas" |
                              site_name == "La Viña"|
                              site_name ==  "Sirgenstein Cave") ~ "1",
                           (site_name == "Bockstein-Törle"|
                              site_name == "Vogelherd" |
                              site_name == "Hohle Fels"|
                              site_name == "Trou Magrite") ~ "2"))

@@

```{r fig.height= 10, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Map of Palaeolithic sites containing signs, sign abundance is represented by point size, data from Dutkiewicz et al. (2020)"
#| 
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)
library(ggpubr)

@@

The modularity test, when run on the above groupings, resulted in a score of `r as_group`. The perMANOVA test produced an R-squared value of `r perman_r` and a P value of `r perman_p`, which tells us that the groupings are statistically significant, and that 52% of the variability in the sign distribution can be attributed to group membership. The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.

@@

plot_grid(trans_strength_plot$, proto_strength_plot, early_strength_plot , evolved_strength_plot, 
          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.03)
```

@@

In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = `r p_value_distance_correlation`, Mantel R = `r r_value_distance_correlation`). Again, the results are displayed in a mantel correlogram (@fig-distance-mantel), which demonstrates that similarity and geographic distance do not experience a clearly correlated relationship.

@@

Through our various analyses, we saw that the isolation-by-distance framework could not account for the distribution of sign types. Specifically, our Mantel test (@fig-distance-mantel) found no statistically significant relationship between geographic distance and sign variation. Therefore, we must look to other factors to explain the distribution. We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and @fig-pcoa. We divided the sites into approximately ten groups based on sign types with a modularity score of `r as_group`. @fig-group-map shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.

>>>>

Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the sub-periods of the Aurignacian period. The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans. Chronology of the Aurignacian is primarly determined by correspondance with those material culture markers, which include carinated scrapers, small-flake tool and split-based bone and ivory tools [@Chu_Richter_2020]. The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k cal BP, and the early Aurignacian at 39.8 - 37.8 k cal BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ended with Greenland Interstadial 8.

```{r}
#| label: fig-aur-table
#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Sanz-Royo_Terlato_Marín-Arroyo_2024, @Chu_Richter_2020; @Fletcher_Sánchez],"
library(knitr)
library(tidyverse)
library(readxl)

aur_table <- read_excel("data/aurignacian-periods.xlsx")

kable(aur_table)

In this study, we aim to address the following questions: How does sign distribution change across phases of the Aurignacian? How do the groupings in sign distribution vary in diversity, size, connectivity, and boundaries? What do the patterns tell us about dynamics of social organization throughout the Aurignacian? First, we identify clusters within each time period using seriation and network analysis. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.

@@

```{r}
## clean the data
library(magrittr)
library(rcarbon)
signbase_full <- read_csv("data/signBase_Version1.0.csv")
nsites <- signbase_full %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts <- nrow(signbase_full)
ncountries <- signbase_full %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes <- ncol(signbase_full)

@@

signbase_full_clean <- signbase_full_clean %>%
  mutate(time_period = case_when((MedianBP > 41499) ~ "transitional",
                                 (MedianBP > 39799) ~ "proto_aurignacian",
                                 (MedianBP > 37899) ~ "early_aurignacian",
                                 (MedianBP > 31999) ~ "evolved_aurignacian"))

@@

trans_unique_data <- trans_unique_data %>% 
  mutate(group = case_when((site_name == "Pod Hradem"|
                              site_name == "Fumane" |
                              site_name == "Labeko Koba") ~ "1",
                           (site_name == "El Castillo" | 
                              site_name == "Hohle Fels" | 
                              site_name == "Geissenklösterle") ~ "2"))

proto_unique_data <- proto_unique_data %>% 
  mutate(group = case_when((site_name == "Gatzarria" | 
                              site_name == "Hohlenstein-Stadel" | 
                              site_name == "Mladeč" | 
                              site_name == "Abri Pataud") ~ "1",
                           (site_name == "Geissenklösterle" |
                              site_name == "Grotte du Renne" |
                              site_name == "Vogelherd" |
                              site_name == "Spy" |
                              site_name == "La Ferrassie" |
                              site_name == "Les Cottés") ~ "2"))

@@

evolved_unique_data <- evolved_unique_data %>% 
  mutate(group = case_when((site_name == "Les Rois" |
                              site_name == "Gargas" |
                              site_name == "La Viña"|
                              site_name ==  "Sirgenstein Cave") ~ "1",
                           (site_name == "Bockstein-Törle"|
                              site_name == "Vogelherd" |
                              site_name == "Hohle Fels"|
                              site_name == "Trou Magrite"|
                              site_name == "Vindija Cave") ~ "2"))

@@

```{r fig.height= 10, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Map of data collection sites, divided by time period, and graph of )"

library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)
library(ggpubr)

@@

The modularity test, when run on the above groupings, resulted in a score of . The perMANOVA test produced an R-squared value of  and a P value of , which tells us that the groupings are statistically significant, and that 52% of the variability in the sign distribution can be attributed to group membership. The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.

@@

@@

In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = , Mantel R = ).

@@

Through our various analyses, we saw that the isolation-by-distance framework could not account for the distribution of sign types. Specifically, our Mantel test () found no statistically significant relationship between geographic distance and sign variation. Therefore, we must look to other factors to explain the distribution. We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and. We divided the sites into approximately ten groups based on sign types with a modularity score of .

shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.
